---
title: PopSV - Bin size comparison
author: Jean Monlong
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, fig.width=10, tidy=TRUE)
```

# PopSV - Bin size comparison

## Load packages, functions and data

```{r}
library(dplyr)
library(magrittr)
library(ggplot2)
library(tidyr)
library(GenomicRanges)
library(BSgenome.Hsapiens.UCSC.hg19)
library(PopSV)
source("EpiPopSV-scripts.R")

ol.df <- function(df, df2){
  df2 = subset(df2, sample==df$sample[1])
  gr2 = with(df2, GRanges(chr, IRanges(start, end)))
  gr = with(df, GRanges(chr, IRanges(start, end)))
  ol = findOverlaps(gr, gr2)
  cov.o = tapply(width(pintersect(gr[queryHits(ol)], gr2[subjectHits(ol)]))/1e3, queryHits(ol), sum)
  df$ol.kb = 0
  df$ol.kb[as.numeric(names(cov.o))] = as.numeric(cov.o)
  df$ol.prop = 1e3 * df$ol.kb / (df$end-df$start+1)
  nb.bin.cons2 = tapply(df2$nb.bin.cons[subjectHits(ol)], queryHits(ol), sum)
  df$nb.bin.cons.2 = 0
  df$nb.bin.cons.2[as.numeric(names(cov.o))] = as.numeric(nb.bin.cons2)
  df
}

## Get normal-tumor information
load("../data/cagekid-5kbp-files.RData")
normals = subset(files.df, status=="normal")$sample

## CNVs from PopSV, FREEC, CNVnator, cn.MOPS, LUMPY
load("../data/cnvs-PopSV-cagekid-5kbp-FDR001.RData")
res.lg = subset(res.df, sample %in% normals)
load("../data/cnvs-PopSV-cagekid-500bp-FDR001.RData")
res.sm = subset(res.df, sample %in% normals)

## Labels
lab.lg="5kb"
lab.sm="500bp"
NB.CORES=3
```

## Control: Random regions with similar sizes

```{r}
seql.1.22 = seqlengths(BSgenome.Hsapiens.UCSC.hg19)[paste0("chr",1:22)]
rand.df <- function(df){
  widths = df$end-df$start+1
  chrs = apply(rmultinom(length(widths), 1, (seql.1.22/1e3)/sum(seql.1.22/1e3)),2,function(e)which(e==1))
  starts = runif(length(widths), 0, seql.1.22[chrs]-widths)
  df$chr = chrs
  df$start = starts
  df$end = starts + widths -1
  df
}
```


## How many small-bin calls overlap large-bin call ?

```{r}
sm.lg = res.lg %>% group_by(sample) %>% do(ol.df(., res.sm))
sm.lg$ol.prop.class = cut(sm.lg$ol.prop, breaks=c(-Inf, 0, .2, .8, 1), labels=c("0","(0,0.2]","(0.2,0.8]","(0.8,1]"))
sm.lg$size = cut(sm.lg$nb.bin.cons, breaks=c(0,1,3,10, Inf))
sm.lg.prop = sm.lg %>% group_by(size, ol.prop.class) %>% summarize(nb=n()) %>% group_by(size) %>% mutate(prop=nb/sum(nb))
ggplot(sm.lg.prop, aes(x=size,fill=ol.prop.class, y=prop)) + ggplot2::geom_bar(colour="black",stat="identity") + theme_bw() +  xlab(paste0(lab.lg," call size (bin)")) + ylab(paste0("proportion of ",lab.lg," calls")) + scale_fill_brewer(name=paste0("proportion of the ",lab.lg," call\n supported by ",lab.sm," calls"))
```

### Significance threshold effect

```{r}
sm.lg.l = res.lg %>% filter(cn2.dev>.3 & qv<1e-4) %>% group_by(sample) %>% do(ol.df(., res.sm))
sm.lg.l$ol.prop.class = cut(sm.lg.l$ol.prop, breaks=c(-Inf, 0, .2, .8, 1), labels=c("0","(0,0.2]","(0.2,0.8]","(0.8,1]"))
sm.lg.l$size = cut(sm.lg.l$nb.bin.cons, breaks=c(0,1,3,10, Inf))
sm.lg.l.prop = sm.lg.l %>% group_by(size, ol.prop.class) %>% summarize(nb=n()) %>% group_by(size) %>% mutate(prop=nb/sum(nb))
ggplot(sm.lg.l.prop, aes(x=size,fill=ol.prop.class, y=prop)) + ggplot2::geom_bar(colour="black",stat="identity") + theme_bw() +  xlab(paste0(lab.lg," call size (bin)")) + ylab(paste0("proportion of ",lab.lg," calls")) + scale_fill_brewer(name=paste0("proportion of the ",lab.lg," call\n supported by ",lab.sm," calls"))
```

### Z-score distribution

This analysis requires the Z-scores for all 500 bp bins in the genome. Because this data is quite big, we only deposited 10 samples, chosen randomly from the 95 normal samples.

```{r, cache=TRUE}
getZsm <- function(df){
  samp = as.character(df$sample[1])
  z.f = paste0("../data/",samp,"-500bp-z.tsv.bgz")
  df$zsm.med = unlist(mclapply(1:nrow(df), function(ii){
    z.call = read.bedix(z.f, df[ii, ])
    median(z.call$z, na.rm=TRUE)
  }, mc.cores=NB.CORES))
  df
}
samples.z = unique(res.lg$sample)
samples.z = samples.z[file.exists(paste0("../data/",samples.z,"-500bp-z.tsv.bgz"))]
sm.lg.z = res.lg %>% filter(sample %in% samples.z) %>% group_by(sample) %>% do(getZsm(.))
sm.lg.z$zsm.med = winsorF(sm.lg.z$zsm.med, 10, -10)
sm.lg.z$z = winsorF(sm.lg.z$z, 20, -20)
```

```{r}
ggplot(subset(sm.lg.z, z>0), aes(x=zsm.med,y=z)) + geom_bin2d(aes(fill=cut(..count.., breaks=c(0,1,10,50,100,1000,Inf)))) + theme_bw() + scale_fill_brewer(name="number of calls") + geom_vline(xintercept=0, linetype=2) + geom_hline(yintercept=0, linetype=2) + theme(legend.position="bottom") + xlab(paste0("median Z-score in ",lab.sm," bins")) + ylab(paste0("Z-score in ",lab.lg," calls")) + scale_x_continuous(breaks=seq(-10,10,2), labels=c(seq(-10,8,2), ">10"), limits=c(-11,11)) + scale_y_continuous(breaks=seq(0,20,5), labels=c(seq(0,15,5), ">20"))
ggplot(subset(sm.lg.z, z<0), aes(x=zsm.med,y=z)) + geom_bin2d(aes(fill=cut(..count.., breaks=c(0,1,10,50,100,1000,Inf)))) + theme_bw() + scale_fill_brewer(name="number of calls") + geom_vline(xintercept=0, linetype=2) + geom_hline(yintercept=0, linetype=2) + theme(legend.position="bottom") + xlab(paste0("median Z-score in ",lab.sm," bins")) + ylab(paste0("Z-score in ",lab.lg," calls")) + scale_x_continuous(breaks=seq(10,-10,-2), labels=c(seq(10,-8,-2), "<-10"), limits=c(-11,11)) + scale_y_continuous(breaks=seq(0,-20,-5), labels=c(seq(0,-15,-5), "<-20"))
```


## Overlap large bins on small bins

```{r}
bin.size.sm = 500
bin.size.lg = 5000
bin.size.diff = bin.size.lg/bin.size.sm

lg.sm = res.sm %>% group_by(sample) %>% do(ol.df(., res.lg))
lg.sm$size2 = cut(lg.sm$nb.bin.cons.2, breaks=c(0,1,2,3,10,Inf), right=FALSE)
lg.sm$size2 = factor(lg.sm$size2, levels=rev(levels(lg.sm$size2)))

lg.sm.null = mclapply(1:3, function(ii)res.sm %>% group_by(sample) %>% do(ol.df(rand.df(.), res.lg)), mc.cores=NB.CORES)
lg.sm.null = do.call(rbind, lg.sm.null)
lg.sm.null = lg.sm.null %>% group_by(nb.bin.cons) %>% summarize(prop0=mean(nb.bin.cons.2!=0)) %>% group_by(nb.bin.cons) %>% summarize(prop0=mean(prop0))

lg.sm$size = lg.sm$nb.bin.cons*bin.size.sm
lg.sm.null$size = lg.sm.null$nb.bin.cons*bin.size.sm

ggplot() + geom_histogram(aes(x=size, fill=size2), data=lg.sm, position="fill") + theme_bw() + xlab(paste0("size of the ",lab.sm,"-bin call")) + scale_fill_brewer(name=paste0("number of consecutive ",lab.lg," bins")) + theme(legend.position="bottom") + ylab(paste0("proportion of ",lab.sm,"-bin calls")) + geom_line(aes(x=size, y=prop0), data=lg.sm.null, linetype=2) + scale_x_continuous(breaks=seq(0,40,bin.size.diff/2)*bin.size.sm, lim=c(0,40*bin.size.sm))

ggplot() + geom_histogram(aes(x=size, fill=size2=="[0,1)", alpha=size2=="[0,1)"), data=lg.sm, position="fill") + theme_bw() + xlab(paste0("size of the ",lab.sm,"-bin call")) + scale_fill_manual(name=paste0("overlaps ",lab.lg," calls"), values=c("steelblue","white"), label=c("TRUE","FALSE")) + theme(legend.position="bottom") + ylab(paste0("proportion overlapping ",lab.lg,"-bin calls")) + geom_line(aes(x=size, y=prop0), data=lg.sm.null, colour="red", linetype=2) + geom_vline(xintercept=seq(0,40,bin.size.diff)*bin.size.sm, linetype=4) + geom_hline(yintercept=1) + guides(fill=FALSE, alpha=FALSE) + scale_x_continuous(breaks=seq(0,40,bin.size.diff/2)*bin.size.sm, lim=c(0,40*bin.size.sm)) + scale_alpha_manual(values=1:0)
```
